---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
    self_contained: no
---


## Loading and preprocessing the data
## Get the feel of the data

```{r readData}
#Let's read data first
activity_data_full=read.csv("activity.csv")

#Let's remove NAs
activity_data=activity_data_full[is.finite(activity_data_full$steps),]

#This is how our data looks
summary(activity_data)
head(activity_data)

```

## What is mean total number of steps taken per day?
```{r prob1}
#Find out the average daily steps
d=sapply(split(activity_data,activity_data$date),function(chunk) sum(chunk$steps,na.rm = T))

#Let's decorate the data a bit
d=data.frame(d)
names(d)=c("Total_Steps")

#And this is how consolidated data looks like
summary(d)
head(d)
```
This is how our daily steps data looks like
```{r}
library(ggplot2)
qplot(Total_Steps,data=d,geom="histogram",group="")
```

Now let's find out the average and median daily steps taken
```{r}
mean_d=data.frame(sapply(split(activity_data,activity_data$date),function(chunk) mean(chunk$steps,na.rm = T)))
median_d=data.frame(sapply(split(activity_data,activity_data$date),function(chunk) median(chunk$steps,na.rm = T)))
names(mean_d)=c("Mean Steps")
names(median_d)=c("Median Steps")

summary(mean_d)
head(mean_d)
summary(median_d)
head(median_d)
```

What's average and median of total steps taken daily?
```{r}

mean_before=mean(d$"Total_Steps")
print("Mean total steps")
print(mean_before)


median_before=median(d$"Total_Steps")
print("Median total steps")
print(median_before)
```

## What is the average daily activity pattern?
```{r prob2}
mean_d=data.frame(sapply(split(activity_data,activity_data$interval),function(chunk) mean(chunk$steps,na.rm = T)))
names(mean_d)=c("Mean_Steps")
summary(mean_d)
head(mean_d)
```

Now let's plot
```{r}
ggplot(data=mean_d,aes(x=as.numeric(rownames(mean_d)),y=Mean_Steps,group=1))+geom_line() + labs(x="Time Interval", y="Mean Steps")

```

Which interval has the max steps on average?
```{r}
max_interval=as.numeric(rownames(mean_d)[which.max(mean_d$Mean_Steps)])
print("Interval with max average steps")
print(max_interval)

```

## Imputing missing values
```{r prob3}

#Let's check how many rows have missing data
print("Number of rows with NA's")
print(sum(!complete.cases(activity_data_full)))
```

### Fill in the missing data with corresponding 5 min interval average
```{r}
#Need to find mean steps with data without NAs
interval_means=as.numeric(mean_d$Mean_Steps)

#Now let's start from the start and fill in the missing data
activity_data_full[is.na(activity_data_full$steps),"steps"]=interval_means[as.factor(activity_data_full[is.na(activity_data_full$steps),"interval"])]

#Let's verify all the missing data is filled in
print("Number of rows with NA's")
print(sum(!complete.cases(activity_data_full)))

#How does the filled in data look like
summary(activity_data_full)
activity_data=activity_data_full
```
Now let's check the daily total steps again
```{r}

d=sapply(split(activity_data,activity_data$date),function(chunk) sum(chunk$steps,na.rm = T))
d=data.frame(d)
names(d)=c("Total_Steps")
summary(d)
head(d)

#Plot the daily steps
qplot(Total_Steps,data=d,geom="histogram",group="")

mean_d=data.frame(sapply(split(activity_data,activity_data$date),function(chunk) mean(chunk$steps,na.rm = T)))
median_d=data.frame(sapply(split(activity_data,activity_data$date),function(chunk) median(chunk$steps,na.rm = T)))
names(mean_d)=c("Mean Steps")
names(median_d)=c("Median Steps")

summary(mean_d)
head(mean_d)
summary(median_d)
head(median_d)

```


What's average and median of total steps taken daily?
```{r}
mean_after=mean(d$"Total_Steps")
print("Mean total steps")
print(mean_after)

median_after=median(d$"Total_Steps")
print("Median total steps")
print(median_after)
```

Let's compare before and after imputing the data
```{r}

print("Mean total steps before = ")
print(mean_before)

print("Mean total steps after=")
print(mean_after)

print("Median total steps before=")
print(median_before)

print("Median total steps after=")
print(median_after)

```

## Are there differences in activity patterns between weekdays and weekends?
```{r prob4}

#Let's find out which days are weekdays and which are weekends
library(lubridate)
dates=ymd(activity_data$date)
wd=wday(dates)
weekendcheck=wd

#Assuming sun and sat are weekends
weekendcheck[!(wd==1 | wd==7)]=0
weekendcheck[weekendcheck!=0]=1
activity_data=cbind(activity_data,factor(x=weekendcheck,labels=c("weekday","weekend")))

#Let's group based on weekend or not and check how the data looks like
library(plyr)
groupcols=c("interval","weekendcheck")
datacols=c("steps")
mean_d=ddply(activity_data,groupcols, function(x) colMeans(x[datacols]))
mean_d$weekendcheck=factor(x=mean_d$weekendcheck,labels=c("weekday","weekend"))
summary(mean_d)
head(mean_d)
```
Let's plot the data grouped by weekend or not
```{r}
ggplot(data=mean_d,aes(x=as.numeric(interval),y=steps,group=1))+geom_line()+facet_grid(weekendcheck~.) + labs(x="Interval",y="No of Steps")

```

That's it, r markdown rocks...!!!